---
title: "Data visualization using ggplot2"
author: "kipngenokoech"
date: "7/10/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning = FALSE)
```


## Introduction
Data visualization is presentation of data and information in pictorial
format.The visual elements like charts ,graphs and maps provides an easy
way of understanding the data you are working with.

some of the advantages of data visualization includes:

  - Better analysis.

  - Quick action.

  - Identifying patterns.

  - Finding errors.

  - Understanding the story.

  - Exploring business insights.

  - Grasping the Latest Trends.

## Reading data and loading packages
``` {r}
library(gridExtra)
library(grid)
library(plyr)
library(tidyverse)
crime<-read.csv("shootings.csv")
```


We are required to convert a number of variables to factors

``` {r}
#function to convert variables to factors
to.factors <- function(df, variables){
  for (variable in variables){
    df[[variable]] <- as.factor(df[[variable]])
  }
  return(df)
}
```

``` {r}
categorical<-c("manner_of_death","armed","gender","race","city","state","signs_of_mental_illness","flee","body_camera","arms_category")
crime<-to.factors(df= crime,variables =categorical)
```

``` {r}
#Extract year from date
crime$date <- as.Date(crime$date)
crime$year <- as.numeric(format(crime$date, "%Y"))
crime$year<-as.factor(crime$year)
crime<-crime[-c(1,3,12)]
```

##data visualization
``` {r}
#checking the random sampling of data
crime[sample(nrow(crime),10),]
```

  

## Data Exploration using GGPLOT

``` {r}
attach(crime)
ggplot(crime,aes(x=age))+geom_histogram()+labs(title = "DISTRIBUTION OF AGE",x="AGE")+theme(panel.background = element_rect(fill = "lightcyan3"),plot.title = element_text(hjust = 0.5,face = "bold",colour = "cadetblue"))+geom_vline(data=crime, aes(xintercept = mean(age)),linetype="dashed",color="deeppink")
```



The distribution of ages is somehow symmetrical but the tail is longer
on the right side which makes it positively skewed.

``` {r}
#checking presence of outliers using boxplot

ggplot(crime,aes(x=race,y=age,fill=signs_of_mental_illness))+geom_boxplot()+theme(panel.background = element_rect(fill = "lightcyan3"),plot.title = element_text(hjust = 0.5,face = "bold",colour = "cadetblue"),legend.background = element_rect(fill="aquamarine"))+labs(title = "BOXPLOT OF AGE W.R.T RACE AND MENTAL ILLNESS ")
```


The victims with mental illness have higher age on average compared to
victims without mental illness.

``` {r}
ggplot(crime,aes(x=race,y=age,fill=race))+geom_violin()+geom_boxplot(width=0.1)+theme(panel.background=element_rect(fill="lightcyan3"),plot.title=element_text(hjust=0.5,colour="cadetblue",face="bold"),legend.background = element_rect(fill="aquamarine"))+labs(title = "DISTRIBUTION OF AGE W.R.T RACE")
```



A combination of boxplot and density plot. The shape of violin shows the
distribution of frequencies.From the plot above  many data points cluster around the median
value.Boxplot depicts the five number summary, maximum,minimum,median,
1st quartile and 3rd quartile.

``` {r}
p1<-ggplot(crime,aes(age,fill=race,colour=race))+geom_density(alpha=.5)+geom_vline(aes(xintercept=mean(age)))+labs(title = "AGE DENSITY PLOT W.R.T race")+theme(panel.background = element_rect(fill="lightcyan3"),plot.title = element_text(face = "bold",colour = "cadetblue",hjust = 0.5),legend.background = element_rect(fill="aquamarine"))
```

``` {r}
p2<-ggplot(crime,aes(age,fill=year,colour=year))+geom_density(alpha=.5)+geom_vline(aes(xintercept=mean(age)))+labs(title = "AGE DENSITY PLOT W.R.T year")+theme(panel.background = element_rect(fill="lightcyan3"),plot.title = element_text(face = "bold",colour = "cadetblue",hjust = 0.5),legend.background = element_rect(fill="aquamarine"))
```

``` {r}
p3<-ggplot(crime,aes(age,fill=body_camera,colour=body_camera))+geom_density(alpha=.5)+geom_vline(aes(xintercept=mean(age)))+labs(title = "AGE DENSITY PLOT W.R.T BODY CAM")+theme(panel.background = element_rect(fill="lightcyan3"),plot.title = element_text(face = "bold",colour = "cadetblue",hjust = 0.5),legend.background = element_rect(fill="aquamarine"))
```

``` {r}
p4<-ggplot(crime,aes(age,fill=manner_of_death,colour=manner_of_death))+geom_density(alpha=.5)+geom_vline(aes(xintercept=mean(age)))+labs(title = "AGE DENSITY PLOT W.R.T MANNER OF DEATH")+theme(panel.background = element_rect(fill="lightcyan3"),plot.title = element_text(face = "bold",colour = "cadetblue",hjust = 0.5),legend.background = element_rect(fill="aquamarine"))
```

``` {r}
grid.arrange(p1 + ggtitle(""),
             p2 + ggtitle(""),
             p3 + ggtitle(""),
             p4 + ggtitle(""),
             nrow = 2,
             top = textGrob("Shootings Density plot", 
                            gp=gpar(fontsize=15)))
```


Density plot is used to show the distribution of one or more variables.From the above we can observe that the distribution of our variables have a long tail towards the right hence positively skewed distributions.